Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Worked on experiments.

As a proxy for genuine graphics rendering, we devised the following
procedure to determine non-cooperative OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgrouops of a
given GPU for a parameterised length of time $t$.  We wrote a host
program to repeatedly launch this kernel in an infinite loop.  We then
searched for a maximum value for the parameter $t$ that allowed the
synthetic kernel to execute without having an observable impact on
graphics rendering, where ``observable'' was decided manually by one
of the paper authors.  Having found the appropriate value for $t$, we
ran the host application for $X$ seconds, measuring the time $Y < X$
dedicated to GPU execution during this period, as well as the number
of kernel launches $n$ that were issued.  We used $X \geq 10$ in all
experiments.  The values $(X-Y)/n$ and $X/n$ estimate the average time
spent using the GPU to render the display between kernel calls and the
rate at which the OS requires the GPU for display rendering,
respectively.

@@

Our results are the following: light workloads needs to run 3ms every
70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
the medium and heavy workloads, the 40ms rate corresponds to graphics
rendering occuring 25 times per second, which coincides with the human
persistence of vision. The 3ms execution duration of both light and
medium configurations indicates that, as expected, GPU computation is
cheaper for basic display rendering compared with rendering that uses
complex shaders. We use these results to guide our evaluation of the
responsiveness of cooperative scheduling in
\mysec\ref{sec:responsiveness}.

@@

We now assess the performance impact of multitasking between a
cooperative and non-cooperative kernel, and on scheduling
responsiveness.  The performance is measured with respect to the
performance of each kernel run in sequence. Responsiveness, i.e., how
fast a non-cooperative kernel can be scheduled, depends not only on
the scheduler implementation, but also on how often the cooperative
kernel offer workgroups to be killed.

@@

\paragraph{Experimental setup}
For a given cooperative kernel and its data set, we launch the kernel
asynchronously and then repeatedly schedule a non-cooperative kernel
that aims to simulate the intensity of a graphics rendering workload,
according to the 3 classes listed in \mysec~\ref{sec:sizingnoncoop}. In
practice, we use a matrix multiplication kernel as the non-cooperative
workload, with matrix dimensions tailored for each GPU to reach the
appropriate execution duration.  We conduct separate runs where we vary
the number of workgroup slots requested by the non-cooperative kernel,
considering the cases where one, a quarter, a half, and all-but-one, of
the total number of workgroup slots is requested.  Moreover, for the
graph algorithms we try both the regular and query barrier
implementations.

@@

To work around compiler crash and compiler hang bugs in Intel's current
OpenCL drivers, we had to apply some semantics-preserving changes to the
code generated by \kernelmerge{} for a few applications configurations.
We could not obtain results for some configurations where the
non-cooperative kernel asks for all but one workgroups (namely bc, bfs
and l-sssp on all chips, and color on Iris), and could not get the game
application to run at all on HD520; these configurations led to
catastrophic issues such as display freezes or blue screens; we also
attribute these to bugs in Intel's current drivers.

% Whenever a cooperative or non-cooperative kernel terminates, we verify
% that it has really performed its computation by comparing its results
% against the expected one. We encountered two kind of issues during the
% experiments: some configurations would lead to a system freeze or crash
% although they run fine on other GPUs. Although we don't claim that our
% programs are all correct, this issue seems to indicate a bug in the
% OpenCL compiler or driver of the problematic GPU. We also have some
% cooperative kernels that returned invalid results when the
% non-cooperative kernel takes all but one of the cooperative kernel
% workgroups, we decided to ignore these results.

@@

\myfiglong\ref{fig:fine-grained-timing} reports, on three
configurations, the gather and execution time of the non-cooperative
kernel w.r.t. the quantity of workgroups allocated to it. A logarithmic
scale is used for time since gather times tend to be much smaller than
execution times. The horizontal grey lines indicates the desired rate
for non-cooperative kernels.

Let's focus on the left figure which illustrates a work stealing
example: when the non-cooperative kernel is given only one workgroup, it
execution is so long that it cannot permit to terminate on time for a
screen refresh. The gather time is very good though, since the scheduler
only needs to collect one workgroup. As more workgroups are allocated to
the non-cooperative kernel, it can benefits from more computational
resources and thus lower its execution time: here the non-cooperative
kernel becomes fast enough with a quarter (resp. half) of available
workgroups for light (resp. heavy) graphics workload. Inversely, the
gather time augments since the scheduler must collect more and more
workgroups.

The middle and right figures, which depict graph algorithms, expose the
same trend: as the number of non-cooperative workgroups grows, the
execution time lowers and the gather time raises.\footnote{The right
figure has only 3 points per line: recall that for bfs there are no runs
with ``all but one'' non-cooperative workgroups}
%
The gather time is higher on the right figure: the rmat22 input graph is
rather wide than deep, so the graph algorithm reaches resizing barriers
less often than for the ecology input of the middle figure for
instance. The scheduler has thus fewer opportunities to collect
workgroups and gather time increases. Nonetheless, scheduling
responsiveness can benefit from the query barrier: when used, this
barrier let the scheduler collects all needed workgroups as soon as they
hit a resizing barrier. As we can see on both the middle and right
figure, the gather time of the query barrier is almost stable w.r.t. the
number of workgroups that needs to be collected.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{heavy.pdf} 
\caption{Blah}\label{fig:blah}
\end{figure}

\paragraph{Performance}
\myfig~\ref{fig:} reports the overhead brought by the scheduling of
non-cooperative kernels over the cooperative kernel execution time
(geometric mean over all applications and inputs on the Iris chip, other
chips look similar\HEComment{We can put the graphs of ther chips in
appendix}), along with the period (aritmetic mean over all applications
and inputs) at which non-cooperative kernels can be scheduled. We show
results for the three workloads listed in
\mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
period graph correspond to the period goals of the workloads: the higher
(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for the
light (resp. medium and heavy) workload.

Non-cooperative kernels running a workload with a single workgroup leads
to almost no overhead, but the period is far too high to meet the needs
of any of our three workloads. For example, a heavy workload averages a
period of 939 ms. As more workgroups are dedicated to non-cooperative
kernels, they execute quickly enough to be scheduled at the expected
period. More concretely, for the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to meet
their goal period (70 and 40 ms resp.). However, a quarter of the
workgroups are not sufficient to meet the goal for the heavy workload
(giving a mean period of 88 ms). Finally, if half of the workgroups are
allocated to the non-cooperative kernel, the heavy workload gets
averages \~10\% over its goal period (mean of 44 ms). If asking for all
but one, the heavy workload can reach its goal period.

Yet, as expected, allocating more non-cooperative workgroups increases
the execution time overhead of the cooperative kernel. Still the heavy
workloads meet their period (by allocating all but one workgroup for
non-cooperative kernels) for an overhead less than 1.4 on average. Light
and medium workloads meet their period with only a negligable overhead
of the cooperative kernel (less than 1.03 slowdown).

@@

\section{Full Set of Graphs for Responsiveness Experiments}

\ADComment{TODO: we need this.  Probably not the right title.}

>>>>

As a proxy for genuine graphics rendering, we devised the following
procedure to determine non-cooperative OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgrouops of a
given GPU for a parameterised length of time $t$.  We wrote a host
program to repeatedly launch this kernel in an infinite loop.  We then
searched for a maximum value for the parameter $t$ that allowed the
synthetic kernel to execute without having an observable impact on
graphics rendering, where ``observable'' was decided manually by one
of the paper authors.  Having found the appropriate value for $t$, we
ran the host application for $X$ seconds, measuring the time $Y < X$
dedicated to GPU execution during this period, as well as the number
of kernel launches $n$ that were issued.  We used $X \geq 10$ in all
experiments.  The values $(X-Y)/n$ and $X/n$ estimate the average time
spent using the GPU to render the display between kernel calls and the
period at which the OS requires the GPU for display rendering,
respectively.

@@

Our results are the following: light workloads needs to run 3ms every
70ms, medium ones 3ms every 40ms, and heavy ones 10ms every 40ms. For
the medium and heavy workloads, the 40ms period corresponds to graphics
rendering occuring 25 times per second, which coincides with the human
persistence of vision. The 3ms execution duration of both light and
medium configurations indicates that, as expected, GPU computation is
cheaper for basic display rendering compared with rendering that uses
complex shaders. We use these results to guide our evaluation of the
responsiveness of cooperative scheduling in
\mysec\ref{sec:responsiveness}.

@@

We now assess the performance impact and responsiveness of
multitasking between a long-running cooperative kernel and a series of
short, non-cooperative kernel launches.  Performance is measured by
comparing the time for multitasked execution with the time taken to
perform all kernel launches in sequence.  Responsiveness, i.e.\ how
fast a non-cooperative kernel can be scheduled, depends not only on
the scheduler implementation, but also on how often the cooperative
kernel offers workgroups to be killed.

@@

\paragraph{Experimental setup}
For a given cooperative kernel and its data set, we launch the kernel
asynchronously and then repeatedly schedule a non-cooperative kernel
that aims to simulate the intensity of a graphics rendering workload,
according to the three classes of workload discussed in \mysec\ref{sec:sizingnoncoop}. In
practice, we use a matrix multiplication kernel as the non-cooperative
workload, with matrix dimensions tailored for each GPU to reach the
appropriate execution duration.  We conduct separate runs where we vary
the number of workgroups requested by the non-cooperative kernel,
considering the cases where one, a quarter, a half, and all-but-one, of
the total number of workgroup slots is requested.  Moreover, for the
graph algorithms we try both the regular and query barrier
implementations.

@@

To work around compiler crash and compiler hang bugs in Intel's current
OpenCL drivers, we had to apply some semantics-preserving changes to the
code generated by \kernelmerge{} for a few applications configurations.
Driver bugs also meant that we could not obtain results for some configurations where the
non-cooperative kernel asks for all-but-one workgroups (namely bc, bfs
and l-sssp on all chips, and color on Iris), and could not get the game
application to run at all on HD520; these configurations led to
severe failures such as machine freezes and blue screens.

% Whenever a cooperative or non-cooperative kernel terminates, we verify
% that it has really performed its computation by comparing its results
% against the expected one. We encountered two kind of issues during the
% experiments: some configurations would lead to a system freeze or crash
% although they run fine on other GPUs. Although we don't claim that our
% programs are all correct, this issue seems to indicate a bug in the
% OpenCL compiler or driver of the problematic GPU. We also have some
% cooperative kernels that returned invalid results when the
% non-cooperative kernel takes all-but-one of the cooperative kernel
% workgroups, we decided to ignore these results.

@@

\myfiglong\ref{fig:fine-grained-timing} reports, on three
configurations, the gather and execution time of the non-cooperative
kernel w.r.t. the quantity of workgroups allocated to it. 
\ADComment{Not clear: are the times for light and heavy averages?} A logarithmic
scale is used for time since gather times tend to be much smaller than
execution times. The horizontal grey lines indicates the desired period
for non-cooperative kernels.  These graphs show a representative sample of our results; the full set of graphs for all configurations is provided in Appendix~\ref{appendix:extraresponsiveness} (supplementary material).

Let us first focus on the left graph, which illustrates a work stealing
example executing on the Iris GPU.  When the non-cooperative kernel is given only one workgroup, its
execution is so long that it cannot complete within the period required for a screen refresh. The gather time is very good though, since the scheduler
only needs to collect one workgroup. As more workgroups are allocated to
the non-cooperative kernel, its execution time benefits from more computational
resources: here the non-cooperative
kernel becomes fast enough with a quarter (resp.\ half) of available
workgroups for light (resp.\ heavy) graphics workload. Inversely, the
gather time increases since the scheduler must collect more and more
workgroups.

The middle and right graphs show results for graph algorithms on the
other two GPUs.  \ADComment{Added this; is it right?}  These
algorithms use barriers, and we experimented with the regular and
query barrier implementations described in \mysec\ref{ }.  The
results shown are for executions that use the more efficient query barrier, except that we also show the gather time associated with the regular barrier for comparison.  The graphs show a similar trend to the left-most graph: as the number of non-cooperative workgroups grows, the
execution time decreases and the gather time increases.\footnote{The right
figure has only 3 points per line: recall that for bfs there are no runs
with ``all-but-one'' non-cooperative workgroups.}
%
The gather time is higher on the right figure: the rmat22 input graph is
rather wide than deep, so the graph algorithm reaches resizing barriers
less often than for the ecology input of the middle figure for
instance. The scheduler thus has fewer opportunities to collect
workgroups and gather time increases. Nonetheless, scheduling
responsiveness can benefit from the query barrier: when used, this
barrier lets the scheduler collect all needed workgroups as soon as they
hit a resizing barrier. As we can see on both the middle and right
figure, the gather time of the query barrier is almost stable w.r.t.\ the
number of workgroups that needs to be collected.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{heavy.pdf} 
\caption{Performance impact of multitasking cooperative and non-cooperative workloads, and the period with which non-cooperative tasks execute}\label{fig:performance}
\end{figure}

\paragraph{Performance} \ADComment{I don't think it's clear how overhead is computed; we need to make it clear that it is a slow-down.  Let's discuss.}
\myfiglong\ref{fig:performance} reports, for the Iris GPU, the overhead brought by the scheduling of
non-cooperative kernels over the cooperative kernel execution time (geometric mean over all applications and inputs), along with the period (aritmetic mean over all applications
and inputs) at which non-cooperative kernels can be scheduled.
The results for the other GPUs are similar and are graphed in Appendix~\ref{appendix:extraperformance} (supplementary material).
 We show
results for the three workloads listed in
\mysec~\ref{sec:sizingnoncoop}. The two gray horizontal lines in the
period graph correspond to the period goals of the workloads: the higher
(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for the
light (resp. medium and heavy) workload.

Co-scheduling non-cooperative kernels that compute their workload using a single workgroup leads to almost no overhead, but the period is far too high to meet the needs
of any of our three workloads. For example, a heavy workload averages a
period of 939 ms. As more workgroups are dedicated to non-cooperative
kernels, they execute quickly enough to be scheduled at the expected
period. More concretely, for the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to meet
their goal period (70 and 40 ms resp.). However, a quarter of the
workgroups are not sufficient to meet the goal for the heavy workload
(giving a mean period of 88 ms). Finally, if half of the workgroups are
allocated to the non-cooperative kernel, the heavy workload gets
averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
%
Yet, as expected, allocating more non-cooperative workgroups increases
the execution time overhead of the cooperative kernel.

Still the heavy
workloads meet their period (by allocating all-but-one workgroup for
non-cooperative kernels), incurring a slow down of less than 1.4$\times$ on average. Light
and medium workloads meet their period with only a negligable overhead
of the cooperative kernel (less than a 1.03$\times$ slowdown on average).

Overall we are encouraged by our experimental findings, which provide a lower bound on potential performance of our coperative kernels model: implementing the model natively, with device-specific runtime support, could only improve performance and responsiveness compared with that of our megakernel-based prototype.

@@

\section{Graphs for Multitasking Experiments}

In \mysec\ref{sec:responsiveness} we showed a selection of graphs to illustrate our
experimental results related to performance and responsiveness.  Here
we present the full set of graphs.

\subsection{Responsiveness graphs}

We present the full set of graphs exemplified by the examples in
\myfig\ref{fig:fine-grained-timing}; for completeness we reproduce those graphs here too.

\ADComment{Tyler: show all the graphs, perhaps grouped by GPU.}

\subsection{Performance graphs}

\myfiglong\ref{fig:performance} presents data on the performance impact and period associated with multitasking cooperative and non-cooperative workloads, for the Iris GPU.

\ADComment{Tyler: say something like ``Corresponding graphs for the XXX and YYY GPUs are shown in \myfigs\ref{ } and~\ref{ }, respectively.'', adding the graphs.}
